A meta-analysis of the impacts of best management practices on nonpoint source pollutant concentration.

### 2024-05-23 frontiers_submission/frontiers_submission_revision_01.tex
working on responses

Results from each database were first filtered to remove duplicates.
After removal of duplicates, each member of the research team (\emph{n} = 4) was assigned a subset of studies to evaluate if they should be included (Table \ref{tab:criteria}).
Each study was reviewed by two team members and differences in opinion were collectively discussed and agreed upon before progressing.
The remaining studies were split among team members for data extraction (Table \ref{tab:datvars}), again with at least two team members reviewing each study.
If data was provided in figures, the data was extracted with the WebPlotDigitizer tool \citep{rohatgiWebPlotDigitizer2022}.
Searches, review, and data extractions were conducted separately for FIB and nutrient related parameters.
See Figure S1 and S2 for ROSES diagram. BMP data from the systematic review is available in \citet{kikoyoTWRINPSBMPDB2024}.

@@

\caption{\label{tab:datvars}Study and effect variables extracted for review.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{25em}}
\toprule
Variable & Description\\
\midrule
Publication Year & Year the study was published\\
Parameter & The specific pollutant measured\\
Runoff source & Dominant source of runoff (crop fields, livestock pasture, commerical, residential)\\
Source type & Major categorization of runoff source: agricultural or urban\\
BMP & BMP evaluated\\
\addlinespace
BMP Classification & BMP description based on NRCS conservation practice standards and EPA BMP fact sheets\\
BMP Category & BMP categorization based on structural or management\\
BMP Subcategory & BMP subcategorization based on pollutant removal processes\\
Study scale & Spatial scale of the study area (lot/field, community, watershed)\\
Location & Location name used in the study description\\
\addlinespace
State & State where the study was conducted\\
Study area & Drainage area in hectares\\
Longitude & Approximated or reported latitude coordinate\\
Latitude & Approximated or reported longitude coordinate\\
Study years & Year or years when data were collected\\
\addlinespace
N control & Number of control measurements\\
N experiment & Number of experiemental measurements\\
X control & Mean concentration for control measurements\\
X experiment & Mean concentration for experiemental measurements\\
SE control & Standard error of control measurements\\
\addlinespace
SE experiment & Standard error of experimental measurements\\
Minimum control & Minimum control measurement\\
Minimum experiment & Minimum experiment measurement\\
Maximum control & Maximum control measurement\\
Maximum experiment & Maximum experiment measurement\\
\addlinespace
SD control & Standard deviation of control measreuments\\
SD experiment & Standard deviation of experimental measurements\\
Units & Units reported by the study\\
Percent reduction & BMP efficiency for studies that only reported efficiency\\
\bottomrule
\end{tabular}
\end{table}

@@

We used the ``rma.mv'' function in the \emph{metafor} R package to fit multilevel random effects regression models with \emph{ROM} as the effect variable \citep{viechtbauerConductingMetaanalysesMetafor2010, rcoreteamLanguageEnvironmentStatistical2023}.
We fit separate models for FIB, total nitrogen (TN), dissolved inorganic nitrogen (DIN), total phosphorus (TP), orthophosphate (PO\textsubscript{4}), and total suspended sediment (TSS).
Our models specified a nested random effects term accounting for heterogeneity between effect sizes from the same study and for heterogeneity between studies.
\emph{ROM} was used as the effect size which required the exclusion of studies that only provided measures of BMP efficiency and not the underlying data used to derive the metric.
A key feature of meta-analysis is the weighting of effects using sampling variance of individual effect sizes.
Fifty-nine percent of 222 effect sizes were missing standard deviations required to estimate sampling variance.
Removal of studies due to missing variance information can introduce substantial bias \citep{kambachConsequencesMultipleImputation2020}.
Missing standard deviations were imputed using the pooled ratio of the mean effect size to coefficient of variation (CV) \citep{brackenStatisticalMethodsAnalysis1992}.
Sampling variance was estimated utilizing the average squared CV across all studies divided by sample size for each effect \citep{nakagawaRobustReadilyImplementable2023, doncasterCorrectionBiasMeta2018}:

@@

where \emph{v} represents the sampling variance, \(CV_{control,i}\) and \(CV^2_{experiment,i}\) are the coefficients of variation from the \emph{i}th study for studies 1, 2, \ldots, \emph{K}. \emph{n\textsubscript{control}} and \emph{n\textsubscript{experiment}} are the number of samples in the control (pre-treatment) trial or experimental (post-treatment) trial respectively.

Our initial models included log transformed influent concentration, BMP subcategory (drainage modification, crop field management, livestock management, filtration, treatment, detention, or infiltration), aridity index (mean-centered), influent concentration\texttimes BMP subcategory interactions, and aridity index\texttimes BMP subcategory interactions were included as fixed effect terms.
Aridity index was the only moderator not obtained directly in the systematic review (Table \ref{tab:datvars}).
We mapped study location coordinates to aridity index values published in the ``Global Aridity Index and Potential Evapotranspiration Database - Version 3'' (Global-AI\_PET\_v3) which provides gridded 30 arc-second annual average precipitation and potential evapotranspiration estimates \citep{zomerVersionGlobalAridity2022}.
The aridity index is calculated as the ratio of mean annual precipitation to mean annual evapotraspiration with values between 0 and 0.5 considered hyper to semi-arid, and values above 0.65 as humid.

@@

Our systematic review identified a total of 33 studies and 125 effect sizes on FIB, 24 studies and 50 effect sizes on TN, 31 studies and 88 effect sizes for DIN, 31 studies and 61 effect sizes for TP, 17 studies and 36 effect sizes for PO\textsubscript{4}, and 33 studies with 125 effect sizes for TSS.
The majority of studies were identified as smaller scaled lot or field studies (Figure \ref{fig:bmpsummary} A).
FIB studies had a roughly equal proportion of large watershed/catchment studies and studies conducted at the community/farm scale or smaller.
We also identified that the majority of studies (all parameters) were conducted on urban or non-agricultural runoff (Figure \ref{fig:bmpsummary} B).
We did identify a wide variety of BMPs in the review, but it did not appear that any particular type of BMP was responsible for the majority of studies for any given parameter (Figure \ref{fig:bmpsummary} C).
Our review was restricted to studies published after between 1999 and 2023.
The number of studies published for each parameter were roughly uniformly distributed over time (Figure \ref{fig:bmptemporal} A) and are not indicative of increases or decreases in the number of published studies.
Study length was strongly skewed for all parameters (Figure \ref{fig:bmptemporal} B). Median study lengths were 3 (DIN), 2 (FIB), 2.5 (PO\textasciitilde4), 2.5 (TN), 2.5 (TP), and 2 (TSS) years.
There appears to be a strong clustering of BMP studies in the mid-Atlantic region (North Carolina, Virginia, Maryland) with other states sparsely represented or completely absent from the review (Figure \ref{fig:studymap}).

@@

Study design prevented us from properly assessing BMP effectiveness as a function of age.
Conducting a meta-analysis of BMP effectiveness over time is hampered both by the lack of long-term studies and lack of standardized reporting mechanisms.
Some studies simply describe the change in pollutant concentrations or loads at the beginning and end of the study as a percent change \citep{haileHydraulicPerformancePollutant2016} which presents statistical problems, especially when sampling variance is not reported.
Changes in performance can also be described using a linear regression using date (transformed as a numeric variable) as an independent variable and log-transformed water quality as the dependent variable \citep{mitschCreatingWetlandsPrimary2012, mitschValidationEcosystemServices2014, pausAssessmentHydraulicToxic2014}.
Slopes are a valid effect size for use in meta-analysis but the set of covariates used between studies should be the same since the coefficient of interest is adjusted to account for other terms in the the regression model \citep{beckerSynthesisRegressionSlopes2007}.
It would be reasonable to assume that regressions equations vary between studies to adjust results for seasonality, flow rates, and other variables.
Future efforts for assessing the performance of BMPs over time would benefit not only from more studies, but a more standardized method for providing comparable results.

@@

Raw data from the systematic review is deposited on Zenodo: \url{https://zenodo.org/doi/10.5281/zenodo.8302472}.
R code used in the analysis is also deposited on Zenodo: TBD.

@@

\begin{figure}
\includegraphics[width=1\linewidth,]{../figures/bmp_summary} \caption{Summary of (A) study scale, (B) dominant runoff source, and (C) BMPs identified in the systematic review.}\label{fig:bmpsummary}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,]{../figures/bmp_temporal_summary} \caption{Number of studies identified in the systematic review summarized by (A) publication date and (B) study length.}\label{fig:bmptemporal}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/study_map} \caption{Distribution of studies identified in the systematic review by state and parameter.}\label{fig:studymap}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/overall_effect} \caption{Estimated effect sizes and intervals from the intercept only multilevel random effects model. Individual points represent studies, with size scaled by sampling variance. The point estimate with uncertaintity bars indicate the estimated overall effect, 95\% confidence intervals, and 95\% prediction intervals. Here, \textit{k} indicates the number of overall effects with the number of unique studies in parenthesis.}\label{fig:overalleffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/bac_predicted} \caption{Predicted marignal effect of influent FIB and aridity index (conditioned on BMP subcategory). Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance.}\label{fig:fibeffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/phos_predicted} \caption{Predicted marignal effect of influent pollutant concentration on TP (Panel A) and PO\textsubscript{4} (Panel B) reductions. Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance.}\label{fig:phoseffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=0.5\linewidth,]{../figures/aridity_density} \caption{Comparison of the relative distributions (density) of aridity index values across the U.S. against aridity index values for studies in the systematic review.}\label{fig:aiindex}
\end{figure}

>>>>

Results from each database were first filtered to remove duplicates.
After removal of duplicates, each member of the research team (\emph{n} = 4) was assigned a subset of studies to evaluate if they should be included (Table \ref{tab:criteria}).
Each study was reviewed by two team members and differences in opinion were collectively discussed and agreed upon before progressing.
The remaining studies were split among team members for data extraction (Table \ref{tab:datvars}), again with at least two team members reviewing each study.
If data was provided in figures, the data was extracted with the WebPlotDigitizer tool \citep{rohatgiWebPlotDigitizer2022}.
Searches, review, and data extractions were conducted separately for FIB and nutrient related parameters.
See Figure S1 and S2 for RepOrting standards for Systematic Evidence Syntheses
(ROSES) diagram.
BMP data from the systematic review is available in \citet{kikoyoTWRINPSBMPDB2024}.

@@

\caption{\label{tab:datvars}Study and effect variables extracted for review.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{25em}}
\toprule
Variable & Description\\
\midrule
Publication Year & Year the study was published\\
Parameter & The specific pollutant measured\\
Runoff source & Dominant source of runoff (crop fields, livestock pasture, commerical, residential)\\
Source type & Major categorization of runoff source: agricultural or urban\\
BMP & BMP evaluated\\
\addlinespace
BMP Classification & BMP description based on NRCS conservation practice standards and EPA BMP fact sheets\\
BMP Category & BMP categorization based on structural or management\\
BMP Subcategory & BMP subcategorization based on pollutant removal processes\\
Study scale & Spatial scale of the study area (lot/field, community, watershed)\\
Location & Location name used in the study description\\
\addlinespace
State & State where the study was conducted\\
Study area & Drainage area in hectares\\
Longitude & Approximated or reported latitude coordinate\\
Latitude & Approximated or reported longitude coordinate\\
Study years & Year or years when data were collected\\
\addlinespace
N control & Number of control measurements\\
N experiment & Number of experiemental measurements\\
X control & Mean concentration for control measurements\\
X experiment & Mean concentration for experiemental measurements\\
SE control & Standard error of control measurements\\
\addlinespace
SE experiment & Standard error of experimental measurements\\
Minimum control & Minimum control measurement\\
Minimum experiment & Minimum experiment measurement\\
Maximum control & Maximum control measurement\\
Maximum experiment & Maximum experiment measurement\\
\addlinespace
SD control & Standard deviation of control measurements\\
SD experiment & Standard deviation of experimental measurements\\
Units & Units reported by the study\\
Percent reduction & BMP efficiency for studies that only reported efficiency\\
\bottomrule
\end{tabular}
\end{table}

@@

We used the ``rma.mv'' function in the \emph{metafor} R package (R version 4.3.1) to fit multilevel random effects regression models with \emph{ROM} as the effect variable \citep{viechtbauerConductingMetaanalysesMetafor2010, rcoreteamLanguageEnvironmentStatistical2023}.
We fit separate models for FIB, total nitrogen (TN), dissolved inorganic nitrogen (DIN), total phosphorus (TP), orthophosphate (PO\textsubscript{4}), and total suspended sediment (TSS).
Our models specified a nested random effects term accounting for heterogeneity between effect sizes from the same study and for heterogeneity between studies.
\emph{ROM} was used as the effect size which required the exclusion of studies that only provided measures of BMP efficiency and not the underlying data used to derive the metric.
A key feature of meta-analysis is the weighting of effects using sampling variance of individual effect sizes.
Fifty-nine percent of 222 effect sizes were missing standard deviations required to estimate sampling variance.
Removal of studies due to missing variance information can introduce substantial bias \citep{kambachConsequencesMultipleImputation2020}.
Missing standard deviations were imputed using the pooled ratio of the mean effect size to coefficient of variation (CV) \citep{brackenStatisticalMethodsAnalysis1992}.
Sampling variance was estimated utilizing the average squared CV across all studies divided by sample size for each effect \citep{nakagawaRobustReadilyImplementable2023, doncasterCorrectionBiasMeta2018}:

@@

where \emph{v} represents the sampling variance, \(CV^2_{control,i}\) and \(CV^2_{experiment,i}\) are the squared coefficients of variation from the \emph{i}th study for studies 1, 2, \ldots, \emph{K}. \emph{n\textsubscript{control}} and \emph{n\textsubscript{experiment}} are the number of samples in the control (pre-treatment) trial or experimental (post-treatment) trial respectively.

Our initial models included log transformed influent concentration, BMP subcategory (drainage modification, crop field management, livestock management, filtration, treatment, detention, or infiltration), aridity index (mean-centered), influent concentration\texttimes BMP subcategory interactions, and aridity index\texttimes BMP subcategory interactions were included as fixed effect terms.
Aridity index was the only moderator not obtained directly in the systematic review (Table \ref{tab:datvars}).
We mapped study location coordinates to aridity index values published in the ``Global Aridity Index and Potential Evapotranspiration Database - Version 3'' (Global-AI\_PET\_v3) which provides gridded 30 arc-second annual average precipitation and potential evapotranspiration estimates \citep{zomerVersionGlobalAridity2022}.
The aridity index is calculated as the ratio of mean annual precipitation to mean annual potential (or reference) evapotranspiration with values between 0 and 0.5 considered hyper to semi-arid, and values above 0.65 as humid.

@@

Our systematic review identified a total of 33 studies and 125 effect sizes on FIB, 24 studies and 50 effect sizes on TN, 31 studies and 88 effect sizes for DIN, 31 studies and 61 effect sizes for TP, 17 studies and 36 effect sizes for PO\textsubscript{4}, and 33 studies with 125 effect sizes for TSS.
The majority of studies were identified as smaller scaled lot or field studies (Figure \ref{fig:bmpsummary} A).
FIB studies had a roughly equal proportion of large watershed/catchment studies and studies conducted at the community/farm scale or smaller.
We also identified that the majority of studies (all parameters) were conducted on urban or non-agricultural runoff (Figure \ref{fig:bmpsummary} B).
We did identify a wide variety of BMPs in the review, but it did not appear that any particular type of BMP was responsible for the majority of studies for any given parameter (Figure \ref{fig:bmpsummary} C).
Our review was restricted to studies published from 2000 through 2022.
The number of studies published for each parameter were roughly uniformly distributed over time (Figure \ref{fig:bmptemporal} A) and are not indicative of increases or decreases in the number of published studies.
Study length was strongly skewed for all parameters (Figure \ref{fig:bmptemporal} B). Median study lengths were 3 (DIN), 2 (FIB), 2.5 (PO\textsubscript{4}), 2.5 (TN), 2.5 (TP), and 2 (TSS) years.
There appears to be a strong clustering of BMP studies in the mid-Atlantic region (North Carolina, Virginia, Maryland) with other states sparsely represented or completely absent from the review (Figure \ref{fig:studymap}).

@@

Study design prevented us from properly assessing BMP effectiveness as a function of age.
Conducting a meta-analysis of BMP effectiveness over time is hampered both by the lack of long-term studies and lack of standardized reporting mechanisms.
Some studies simply describe the change in pollutant concentrations or loads at the beginning and end of the study as a percent change \citep{haileHydraulicPerformancePollutant2016} which presents statistical problems, especially when sampling variance is not reported.
Changes in performance can also be described using a linear regression using date (transformed as a numeric variable) as an independent variable and log-transformed water quality as the dependent variable \citep{mitschCreatingWetlandsPrimary2012, mitschValidationEcosystemServices2014, pausAssessmentHydraulicToxic2014}.
Slopes are a valid effect size for use in meta-analysis but the set of covariates used between studies should be the same since the coefficient of interest is adjusted to account for other terms in the regression model \citep{beckerSynthesisRegressionSlopes2007}.
It would be reasonable to assume that regressions equations vary between studies to adjust results for seasonality, flow rates, and other variables.
Future efforts for assessing the performance of BMPs over time would benefit not only from more studies, but a more standardized method for providing comparable results.

@@

Raw data from the systematic review is deposited on Zenodo: \url{https://zenodo.org/doi/10.5281/zenodo.8302472}.
R code used in the analysis is also deposited on Zenodo: \url{https://zenodo.org/records/10795231}.

@@

\begin{figure}
\includegraphics[width=1\linewidth,]{../figures/bmp_summary} \caption{Summary of (A) study scale and (B) dominant runoff source. Water quality parameters include total suspended sediment (TSS), total phosphorus (TP), total nitrogen (TN), orthophosphate (PO\textsubscript{4}), fecal indicator bacteria (FIB), and dissolved inorganic nitrogen (DIN).}\label{fig:bmpsummary}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,]{../figures/bmp_summary_param} \caption{Summary BMPs identified in the systematic review by tested parameter. Water quality parameters include total suspended sediment (TSS), total phosphorus (TP), total nitrogen (TN), orthophosphate (PO\textsubscript{4}), fecal indicator bacteria (FIB), and dissolved inorganic nitrogen (DIN).}\label{fig:bmpsummaryparam}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,]{../figures/bmp_temporal_summary} \caption{Number of studies identified in the systematic review summarized by (A) publication date and (B) study length. Water quality parameters include total suspended sediment (TSS), total phosphorus (TP), total nitrogen (TN), orthophosphate (PO\textsubscript{4}), fecal indicator bacteria (FIB), and dissolved inorganic nitrogen (DIN).}\label{fig:bmptemporal}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/study_map} \caption{Distribution of studies identified in the systematic review by state and parameter. Water quality parameters include total suspended sediment (TSS), total phosphorus (TP), total nitrogen (TN), orthophosphate (PO\textsubscript{4}), fecal indicator bacteria (FIB), and dissolved inorganic nitrogen (DIN).}\label{fig:studymap}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/overall_effect} \caption{Estimated effect sizes and intervals from the intercept only multilevel random effects model. Individual points represent studies, with size scaled by sampling variance. The point estimate with uncertaintity bars indicate the estimated overall effect, 95\% confidence intervals, and 95\% prediction intervals. Here, \textit{k} indicates the number of overall effects with the number of unique studies in parenthesis. Water quality parameters include total suspended sediment (TSS), total phosphorus (TP), total nitrogen (TN), orthophosphate (PO\textsubscript{4}), fecal indicator bacteria (FIB), and dissolved inorganic nitrogen (DIN).}\label{fig:overalleffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/bac_predicted} \caption{Predicted marginal effect of influent fecal indicator bacteria (FIB; Panel A) and aridity index (conditioned on BMP subcategory; Panel B). Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance. Higher aridity index (mean annual precipition/evapotranspiration) indicates more humid conditions.}\label{fig:fibeffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\linewidth,]{../figures/phos_predicted} \caption{Predicted marignal effect of influent pollutant concentration on total phosphorus (TP; Panel A) and orthophosphate (PO\textsubscript{4}; Panel B) reductions. Solid lines are the predicted mean effect, dashed lines are the 95\% confidence intervals, and the dotted lines are the 95\% prediction intervals. Individual dots represent each effect size identified in the literature with the size scaled by sampling variance.}\label{fig:phoseffect}
\end{figure}

\begin{figure}[p]
\includegraphics[width=0.5\linewidth,]{../figures/aridity_density} \caption{Comparison of the relative distributions (density) of aridity index values (mean annual precipition/evapotranspiration) across the U.S., across U.S. agricultural and developed land uses, and for BMP study locations. Higher aridity index values indicate more humid conditions.}\label{fig:aiindex}
\end{figure}
